---
title: "eds222_final_project"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
set.seed(67)
```

```{r}
# reading in data
grape_data <- read.csv("GRAPE_QUALITY.csv")
```

```{r}
# viewing a summary of the type of columns and values
str(grape_data)
```
```{r}
# checking for NAs
which(is.na(grape_data))
```
```{r}
# summarizing columns
summary(grape_data)
```
# Visualization

```{r}
# graphing and color coding the quality of grapes
ggplot(grape_data, aes(rainfall_mm, sugar_content_brix, colour = quality_category)) +
  geom_point()

```
```{r}
# graphing and color coding the quality of grapes
ggplot(grape_data, aes(sun_exposure_hours, sugar_content_brix, colour = quality_category)) +
  geom_point()

```


```{r}
# graphing and color coding the quality of grapes
ggplot(grape_data, aes(rainfall_mm, quality_score, colour = quality_category)) +
  geom_point()
```
## We can see that scores of quality of grapes are about a half point each. We can see visually that the quality of grapes MAY go down as rainfall is low of high.


```{r}
# graphing the quality score against sun exposure
ggplot(grape_data, aes(sun_exposure_hours, quality_score, colour = quality_category)) +
  geom_point()
```
# The graph depicts the positive increasing relationship, however, interestingly enough it seems that for each type of grape, incrasing sun exposore have a decreasing effect of measured grapes. This leads to believe that there is a sweet spot for sun exposure. Applying an analysis similar to pesticide abundance might prove useful.
## Interesting points
  - plot would be related to the ribbon and CI of means of scores
  - Based on the data, it seems that only high quality grapes tend to need 9+ hours of sun exposure
  - For low quality grapes, the number of these grapes tend to decrease as 

```{r}
ggplot(grape_data, aes(quality_category)) +
  geom_bar()
```
# Based on this, we have a majority of grapes that are within the medium and high quality category. This may be due to just lack of data or a causal effect of variables 

```{r}
grape_data %>%
  ggplot (aes(berry_size_mm)) +
  geom_density(fill = "darkgreen")
```
```{r}
  ggplot(grape_data, aes(x = quality_category, y = quality_score)) +
  geom_boxplot()
```
## Based on this graph, we can see that 

```{r}
ggplot(grape_data, aes(x = region, y = quality_score)) +
  geom_boxplot()
```

